---
title: "NBA MVP Predictions"
author: "Ian Bogley, Jonas Bowman"
date: "2/22/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning = FALSE)
```

A constant in the sporting world is attempting to predict various metrics associated with these unpredictable games. From march madness, where individuals make their bets for how college playoffs will turn out to sports fans arguing over whether their team will succeed.

An interesting application of this may be in the MVP and All-star selections, where players are chosen based off of voting systems associated with media spectators and individual ones. One question is whether we might be able to use computer algorithms and data analysis to predict how voters chose these prestigious awards.

Let's use standard machine learning techniques to attempt predictions on an NBA season's MVP and all-star selections. To start, lets load the packages we need for this project.

```{r intro}
###PACKAGES###
library(pacman)
p_load(rvest,tidyverse,janitor)


###DATA###
##Read in data, use janitor::clean_names for renaming columns

#background info
playerstats_df <- read.csv("data/player_data.csv") %>% clean_names()%>%rename(player = name)

#physical characteristics
players_df <- read.csv("data/Players.csv") %>% clean_names() %>%
  mutate(player = gsub(" +$","",gsub("[^[:alpha:]]"," ",player)))

#season stats
seasonstats_df <- read.csv("data/Seasons_Stats.csv") %>%
  #Change year format to be more specific (from xxxx-yy to xxxx-yyyy)
  mutate(season = paste(Year-1,Year,sep = "-")) %>%
  #reorder columns for convenience
  .[,c(1,2,54,3:53)] %>% clean_names() %>%
  #Only considering data after the 1985 season
  filter(year>1985) %>%
  #Clean up player names
  mutate(player = gsub(" +$","",gsub("[^[:alpha:]]"," ",player)))

#Read in wikipedia scrapped data
#MVP data
mvp <- read_html("https://en.wikipedia.org/wiki/NBA_Most_Valuable_Player_Award") %>%
  html_nodes(xpath = "//*[@id='mw-content-text']/div[1]/table[4]") %>%
  html_table() %>% data.frame() %>% clean_names() %>%
  #clean up season and player variables
  mutate(season = gsub("[^[:alnum:]]","-19",season),
         player = gsub(" +$","",gsub("[^[:alpha:]]"," ",player))) %>%
  #rename player variable to ensure smooth joining with other dfs
  rename("year_mvp" = "player") %>%
  .[,c(1,2)]
mvp$season[45:65] <- gsub("^([0-9]{4})([^[:alnum:]]{1}[0-9]{2})([0-9]{2})$","\\1-20\\3",mvp$season[45:65])
mvp$year_mvp[44] <- "Karl Malone"
mvp$year_mvp[56] <- "Derrick Rose"

#All-star selections
allstars <- read_html("https://en.wikipedia.org/wiki/List_of_NBA_All-Stars") %>%
  html_nodes(xpath = "//*[@id='mw-content-text']/div[1]/table[2]") %>%
  html_table() %>% data.frame() %>% clean_names() %>%
  #clean up player variable, and prepare allstar selections for future use 
  mutate(player = gsub(" +$","",gsub("[^[:alpha:]]"," ",player)),
         selections = strsplit(selections_c,split = ";"),
         selections_c = NULL) %>%
  .[,c(1,5)]
#manually fix minor issues from webscraping
allstars$player[1] <- "Kareem Abdul Jabbar"
allstars$player[19] <- "Hakeem Olajuwon"
allstars$player[430] <- "Metta World Peace"
```


```{r intro}
#merging the player df's
df_halffull <- left_join(seasonstats_df, players_df,playerstats_df, by = "player")

#create mvp column with TRUE/FALSE values
full_df <- left_join(df_halffull, mvp, by ="season") %>%
  mutate(mvp = (year_mvp == player))

#Create allstar column with TRUE/FALSE values
full_df$allstar <- FALSE
for (i in 1:nrow(allstars)) {
  
  #Pickup selections in their source format from webscrapped table
  selections <- allstars$selections[[i]]
  
  #Turn the format into yyyy:yyyy for further use
  selections_seq <- gsub("[^[:alnum:]]",":",gsub(" ","",selections))
  
  #put years selected into a useable format: a df
  #player= player selected; selections = years selected for allstar game
  years_selected <- data.frame(
    player = allstars$player[i],
    selections = unlist(lapply(selections_seq,function(x) {eval(parse(text = x))}))
  )
  
  #turn values of full_df's allstar column to true based on membership in years_selected
  full_df[(full_df$player == years_selected$player[1] & 
             full_df$year %in% years_selected$selections),"allstar"] <- TRUE
}

#create final df with finalized variables
final_df <- full_df %>%
  #create an age column
  mutate(age = year-born) %>%
  #reorder columns for convenience
  .[,c(2:4,63:64,56:59,60,61,5:54)] %>%
  select(-c(blanl,blank2))
```
NOTE: There wasn't an all-star game in the 1998-1999 season, and there were fewer games due to a lockout.
